---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Ana Andino"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2, message=FALSE, warning = FALSE}
#1

#install.packages(c("tidyverse", "agricolae", "lubridate", "here"))

# Load the packages
library(tidyverse)   
library(agricolae)
library(lubridate)  
library(here)    
library(ggplot2)

# Check your working directory
getwd()

#upload raw data
LakeChemistryPhysics <- read.csv(
  here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
  stringsAsFactors = TRUE)

# Set date to date format
LakeChemistryPhysics$sampledate <- as.Date(LakeChemistryPhysics$sampledate,
                                           format = "%m/%d/%y")

#2

mytheme <- theme_minimal(base_size = 14) +
  theme(axis.text = element_text(color = "grey", face = "bold"),
         axis.title = element_text(size = 12),
                                 legend.position= "bottom")
theme_set(mytheme)
  


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temeprature in July is the same across all lakes (it does not change) across different depths. 
Ha: Mean lake temperature in July changes across different depths. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot, message=FALSE, warning=FALSE}
#4

Lake.subset <- LakeChemistryPhysics %>%
  filter(month(sampledate) == 7) %>%
  select(lakename:daynum, depth:temperature_C) %>%
  drop_na()

#5

plot1 <- ggplot(Lake.subset,
  aes(x= depth, y= temperature_C))+
    geom_point(alpha = 0.5, size = 2, color = "lightblue")+
    geom_smooth(method = lm, se = FALSE, color = "black", size = 0.8)+
  ylim(0 , 35)+
   labs(
    title = "Temperature by Depth in July",
    x = "Depth (m)",
    y = "Temperature (째C)")+
  mytheme

print(plot1)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: There is a negative relationship between depth and temperature which means that as depth increases, temperature decreases. At shallower depths (0-5 meters), the temperature values show greater variability which might indicate there are other variables that might be affecting the variability - that influence  influence temperature near the surface which a linear model is not capturing. At deeper depths you see that variability is smaller. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression, echo=TRUE, warning=FALSE}
#7

reg.depth <- lm(data = Lake.subset, temperature_C ~ depth)

summary(reg.depth)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to the results, 73.05 of the variability in termrpature is explained by changes in depth. The degrees of freedon are 9,726 (n-parameters) and thee results are statistically significat at a 5% level (0.05 alpha) given that the p-values are very small for the intercept and depth but also for the general model (p-value: < 2.2e-16). Thus, we can reject the null hypothesis that depth has no effect on temperature. For every 1 m increase in depth, temperature will decrease by approximately 1.95째C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model, warning=FALSE}
#9

LakesDepth.all <- lm(data = Lake.subset, temperature_C ~ depth + year4 + daynum)
step(LakesDepth.all)


#10

# Multiple regression with the selected predictors
LakesDepth.model <- lm(temperature_C ~ depth + year4 + daynum, data = Lake.subset)
summary(LakesDepth.model)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The stepwise AIC process suggests that all three variables (depth, year4, daynum) should remain in the model, as removing any of them increases the AIC, indicating a worse fit. Approximately 74.11% of the variability is explaines by the independent variables (Xs) which shows a slight improvement to the model. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model, warning=FALSE}
#12

#Using aov
LakeTemp.anova <- aov(data = Lake.subset, temperature_C ~ lakename)
summary(LakeTemp.anova)
#P-values very small thus reject the null hypothesis

# Using Linear model 
LakeTemp.LmAnova <- lm(data = Lake.subset, temperature_C ~ lakename)
summary(LakeTemp.LmAnova)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
There is a significant difference. In both models p-values are very small and almost zero which mans that we reject the null hypothesis (temperatures are the same) and conclude that the lake's average temperature differs from others. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, warning=FALSE, fig.align='center', fig.width=8, fig.height=5}
#14.
plot.template <- ggplot(Lake.subset,
  aes(x= depth, y = temperature_C, color = lakename))+
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se= FALSE)+
    ylim (0, 35)+
    labs(title = "Temperature by Depth in July for Different Lakes",
    x = "Depth (m)",
    y = "Temperature (째C)", color = "Lake Name", size = 3)+
  theme_minimal(base_size = 12) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.title = element_text(face = "plain"),  
    legend.position = "right",
     legend.text = element_text(size = 10), 
    legend.key.width = unit(1.5, "cm")
  ) 

print(plot.template)



```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test, width= 4, lenght = 6}
#15

#still just looking at July 
tukey.anova<- TukeyHSD(LakeTemp.anova)
print(tukey.anova)
#For my own visualization
#plot(tukey.anova, las = 1)

#Finding ways to do it more automatically
tukey_df <- as.data.frame(tukey.anova$lakename)
peter_lake_comparisons <- tukey_df[grep("Peter Lake", rownames(tukey_df)), ]
same_mean_as_peter <- peter_lake_comparisons %>%
  filter(`p adj` > 0.05 & lwr < 0 & upr > 0)
print(same_mean_as_peter)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Pual Lake and Ward Lake have a mean temperature that is statistically the same as Peter Lake and there is no lake that has a statistically distinct mean temperature from all others,

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 
>Answer: Possibly a two sample t- test

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test, warning=FALSE}

Crampton_Ward_data <- Lake.subset %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

# Perform the two-sample t-test
two_test <- t.test(
  temperature_C ~ lakename, 
  data = Crampton_Ward_data)

print(two_test)

```

>Answer: The result matches the result for part 16 thus, the mean temperatures of Crampton Lake and Ward Lake are not significantly different.
